---
title: "p8105_hw2_pt2632"
author: "Pei Tian, pt2632"
date: "10/3/2023"
output: 
  github_document: 
  html_document:
    toc: True
    float_toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, message=FALSE}
library(tidyverse)
```

## Problem 1
  
### Raw Datasets Processing
1. `pols-month.csv`: 822 observations, 9 variables (Before processing)
  
    Variables:
  
    - `mon`: date of the count
  
    - `prez_gop`: indicator of whether the president was republican on the associated date (1 = yes, 0 = no)
  
    - `gov_gop`: the number of republican governors on the associated date
  
    - `sen_gop`: the number of republican senators on the associated date
  
    - `rep_gop`: the number of republican representatives on the associated date
  
    - `prez_dem`: indicator of whether the president was democratic on the associated date (1 = yes, 0 = no)
  
    - `gov_dem`: the number of democratic governors on the associated date
  
    - `sen_dem`: the number of democratic senators on the associated date
  
    - `rep_dem`: the number of democratic representatives on the associated date
  
```{r}
# Tidy 'pols-month.csv'
pols <- 
  # load data
  read_csv("./datasets/HW2/fivethirtyeight_datasets/pols-month.csv") |> 
  # clean column name
  janitor::clean_names() |> 
  # separate date into 'year', 'month', 'day'
  separate(mon, c("year", "month","day"), convert = TRUE, sep = '-') |> 
  # change month number to month name
  mutate(month = month(month, label = TRUE), 
         # create 'president' variable according to value of 'prez_gop' variable
         president = recode(
           prez_gop, 
           "0" = "dem", 
           "1" = "gop",
           "2" = "gop"
         )) |>
  # remove some columns
  select(-c(day, prez_dem, prez_gop)) 
```
After processing, there are `r nrow(pols)` observations and `r ncol(pols)` variables in the dataset. 

The range of the `year` is between `r range(pull(pols, year))[1]` and `r range(pull(pols, year))[2]`.


2. `snp.csv`: 787 rows, 2 columns (Before processing)
  
    Variables:
  
    - `date`: the date of the observation
  
    - `close`: the closing values of the S&P stock index on the associated date
  
```{r}
# Tidy 'snp.csv'
snp <- 
  # load data
  read_csv("./datasets/HW2/fivethirtyeight_datasets/snp.csv") |> 
  # clean column name
  janitor::clean_names() |>  
  # separate date into 'year', 'month', 'day'
  separate(date, c("year", "month","day"), convert = TRUE, sep = '/') |>
  # change month number to month name
  mutate(month = month(month, label = TRUE)) |> 
  # arrange columns
  arrange(year, month) |> 
  # remove some columns
  select(-day)
```
After processing, there are `r nrow(snp)` observations and `r ncol(snp)` variables in the dataset.

The range of the `year` is between `r range(pull(snp, year))[1]` and `r range(pull(snp, year))[2]`.


3. `unemployment.csv`: 816 rows, 13 columns (Before processing)
  
    Variables: 
    - `year`: year information 
  
    - Others: unemployment rate of exact month
  
```{r}
# Tidy 'unemployment.csv'
unemploy <- 
  # load data
  read.csv("./datasets/HW2/fivethirtyeight_datasets/unemployment.csv") |>
  # clean column name
  janitor::clean_names()  |>
  # switch "wide" to "long" format
  pivot_longer(-year, names_to = "month", values_to = "unemploy_rate") |> 
  # capitalize first letter to ensure same values in key variable
  mutate(month = str_to_title(month))  
```
After processing, there are `r nrow(unemploy)` observations and `r ncol(unemploy)` variables in the dataset. 

The range of the `year` is between `r range(pull(unemploy, year))[1]` and `r range(pull(unemploy, year))[2]`.


### Data Merge
**Key variables:** use `year` and `month` variables to merge datasets.
```{r}
# Join datasets (Merge)
res <- left_join(pols, snp, by = join_by(month == month, year == year))
res <- left_join(res, unemploy, by = join_by(month == month, year == year))
```


## Problem 2
### Raw Datasets Processing
1. `Mr.Trash Wheel` sheet:
```{r}
# Tidy "Mr.Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash1 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 1, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
```
After processing, there are `r nrow(trash1)` observations and `r ncol(trash1)` variables in the dataset.

Variables:

  - `dumpster`: dumpster id

  - `month`, `year`, `date`: date information

  - `weight_tons`: collected trash weight

  - `volume_cubic_yards`: trash volume

  - `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `grocery_bags`, `chip_bags`, `sports_balls`: specific trash amount

  - `homes_powered`: power amount 
  
2. `Professor Trash Wheel` sheet:
```{r}
# Tidy "Professor Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash2 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 2, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
```
After processing, there are `r nrow(trash2)` observations and `r ncol(trash2)` variables in the dataset.

Variables:

  - `dumpster`: dumpster id

  - `month`, `year`, `date`: date information

  - `weight_tons`: collected trash weight

  - `volume_cubic_yards`: trash volume

  - `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `grocery_bags`, `chip_bags`: specific trash amount

  - `homes_powered`: power amount 

3. `Gwynnda Trash Wheel` sheet: 
```{r}
# Tidy "Gwynnda Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash3 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 3, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
```
After processing, there are `r nrow(trash3)` observations and `r ncol(trash3)` variables in the dataset.

Variables:

  - `dumpster`: dumpster id

  - `month`, `year`, `date`: date information

  - `weight_tons`: collected trash weight

  - `volume_cubic_yards`: trash volume

  - `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `plastic_bags`: specific trash amount

  - `homes_powered`: power amount 

### Data Combination
```{r}
# combine datasets into single dataset
total_trash <- bind_rows(
  list("Mr. Trash Wheel" = trash1, 
       "Professor Trash Wheel" = trash2, 
       "Gwynnda Trash Wheel" = trash3), .id = "source")

```
After row binding process, there are `r nrow(total_trash)` observations and `r ncol(total_trash)` variables in the dataset. 

Specific questions:  

The total weight of trash collected by Professor Trash Wheel equals to `r sum(pull(trash2, weight_tons))`. 

The total number of cigarette butts collected by Gwynnda in July of 2021
equals to `r trash3 |> filter(month == "July" & year == 2021) |> pull(cigarette_butts) |> sum()`.


## Problem 3
### Demographic Dataset
```{r}
baseline <- 
  # load data
  read_csv("./datasets/HW2/data_mci/MCI_baseline.csv", skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # 'sex' and 'apoe4' variable as string
  mutate(sex = recode(sex, "0" = "Female", "1" = "Male"), 
         apoe4 = recode(apoe4, "0" = "non-carrier", "1" = "carrier"))
```
There are `r nrow(baseline)` observations and `r ncol(baseline)` variables in this dataset.

There are `r dim(baseline)[1]` participants are recruited.

The average age of baseline is `r mean(pull(baseline, current_age))`.

There are `r dim(filter(baseline, apoe4 == "carrier" & sex == "Female"))[1] / dim(filter(baseline, sex == "Female"))[1] * 100`% of women in the study are APOE4 carriers. 


```{r}
# remove participants with no MCI
mci_baseline <- filter(baseline, age_at_onset != ".")
mci_baseline
```
After processing, there are `r nrow(mci_baseline)` observations and `r ncol(mci_baseline)` variables. 

And there are `r dim(mci_baseline)[1]` participants developed MCI.


### Biomarker Dataset
```{r}
amyloid <- 
  # load data
  read_csv("./datasets/HW2/data_mci/mci_amyloid.csv", skip = 1) |>
  # clean column name
  janitor::clean_names()  
```
There are `r nrow(amyloid)` observations and `r ncol(amyloid)` variables in this dataset.


### Data Combination
```{r}
# check whether some participants appear in only the baseline or amyloid datasets
both_id <- intersect(pull(baseline, id), pull(amyloid, study_id))
# baseline
only_baseline <- c()
for(id in pull(baseline, id)){
  if (!id %in% both_id)
    only_baseline <- append(only_baseline, id)
}
# amyloid
only_amyloid <- c()
for(id in pull(amyloid, study_id)){
  if (!id %in% both_id)
    only_amyloid <- append(only_amyloid, id)
}
print("Participant only in baseline:")
only_baseline
print("Participant only in amyloid:")
only_amyloid
```
There are `r length(only_baseline)` participants only in baseline dataset, and `r length(only_amyloid)` participants only in amyloid dataset.

```{r}
# combine datasets
participants <- inner_join(baseline, amyloid, by = join_by(id == study_id))
participants
```
After combining datasets with `id`, the result are of `r nrow(participants)` observations and `r ncol(participants)` variables.

```{r}
# write combined data as csv file
write_csv(participants, "./output/hw2_problem3_output.csv")
```

