---
title: "p8105_hw2_pt2632"
author: "Pei Tian, pt2632"
date: "9/26/2023"
output: 
  github_document: 
  html_document:
    toc: True
    float_toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE}
library(tidyverse)
```

## Problem 1
  
### Raw Datasets Processing
```{r}
# Tidy 'pols-month.csv'
pols <- 
  # load data
  read_csv("./datasets/HW2/fivethirtyeight_datasets/pols-month.csv") |> 
  # clean column name
  janitor::clean_names() |> 
  # separate date into 'year', 'month', 'day'
  separate(mon, c("year", "month","day"), convert = TRUE, sep = '-') |> 
  # change month number to month name
  mutate(month = month(month, label = TRUE), 
         # create 'president' variable according to value of 'prez_dem' variable
         president = recode(
           prez_gop, 
           "0" = "dem", 
           "1" = "gop",
           "2" = "gop"
         )) |>
  # remove some columns
  select(-c(day, prez_dem, prez_gop)) 
pols
```
```{r}
# Tidy 'snp.csv'
snp <- 
  # load data
  read_csv("./datasets/HW2/fivethirtyeight_datasets/snp.csv") |> 
  # clean column name
  janitor::clean_names() |>  
  # separate date into 'year', 'month', 'day'
  separate(date, c("year", "month","day"), convert = TRUE, sep = '/') |>
  # change month number to month name
  mutate(month = month(month, label = TRUE)) |> 
  # arrange columns
  arrange(year, month) |> 
  # remove some columns
  select(-day)
snp
```
```{r}
# Tidy 'unemployment.csv'
unemploy <- 
  # load data
  read.csv("./datasets/HW2/fivethirtyeight_datasets/unemployment.csv") |> 
  # clean column name
  janitor::clean_names()  |>
  # switch "wide" to "long" format
  pivot_longer(-year, names_to = "month", values_to = "unemploy_rate") |> 
  # capitalize first letter to ensure same values in key variable
  mutate(month = str_to_title(month))  
unemploy
```
### Data Merge
**Key variables: **`year` and `month`
```{r}
# Join datasets (Merge)
res <- left_join(pols, snp, by = join_by(month == month, year == year))
res <- left_join(res, unemploy, by = join_by(month == month, year == year))
res
```

### Result Dataset (Merged Dataset)
```{r}
summary(res)
```
```{r}
# dimension of result
dim(res)
```
```{r}
# range of year
range(pull(res, year))
```

### Datasets Description
1. `pols-month.csv`: 822 rows, 9 columns
  Variables:
  - `mon`: 
  - `prez_gop`:
  - `gov_gop`: 
  - `sen_gop`: 
  - `rep_gop`: 
  - `prez_dem`: 
  - `gov_dem`: 
  - `sen_dem`: 
  - `rep_dem`:
2. `snp.csv`: 787 rows, 2 columns
  Variables:
  - `date`: 
  - `close`: 
3. `unemployment.csv`: 68 rows, 13 columns
  Variables: 
  - `year`: date type, year information 
  - Others: unemployment rate of exact month
4. Combined data:
  - 

## Problem 2
### Raw Datasets Processing
```{r}
# Tidy "Mr.Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash1 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 1, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
trash1
```

```{r}
# Tidy "Professor Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash2 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 2, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
trash2
```

```{r}
# Tidy "Gwynnda Trash Wheel" sheet in "Trash Wheel Collection Data.xlsx"
trash3 <- 
  # load data, specify sheet + omit row with figure
  readxl::read_excel("./datasets/HW2/Trash Wheel Collection Data.xlsx", sheet = 3, skip = 1) |> 
  # clean column name
  janitor::clean_names() |>
  # omit columns with notes
  janitor::remove_empty(which = "cols") |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  # calculate 'houses_powered' variable, houses_powered = 500 * weight_tons + 30
  mutate(homes_powered = 500 * weight_tons + 30,
         # convert type of year into 'character'
         year = as.character(year))
trash3
```

### Data Combination
```{r}
# combine datasets into single dataset
total_trash <- bind_rows(list("Mr. Trash Wheel" = trash1, "Professor Trash Wheel" = trash2, "Gwynnda Trash Wheel" = trash3), .id = "source")
total_trash
```

### Some Questions
```{r}
# total weight of trash collected by Professor Trash Wheel

sum(pull(trash2, weight_tons))
```
```{r}
# total number of cigarette butts collected by Gwynnda in July of 2021
trash3 |> 
  filter(month == "July" & year == 2021) |>
  pull(cigarette_butts) |>
  sum()
```



## Problem 3
### Demographic Dataset
```{r}
baseline <- 
  # load data
  read_csv("./datasets/HW2/data_mci/MCI_baseline.csv", skip = 1) |>  
  # clean column name
  janitor::clean_names() |>
  # 'sex' and 'apoe4' variable as string
  mutate(sex = case_match(sex, 
                          0 ~ "Female",
                          1 ~ "Male"), 
         apoe4 = case_match(apoe4, 
                            0 ~ "non-carrier",
                            1 ~ "carrier"))
```
```{r}
# remove participants with no MCI
mci_baseline <- filter(baseline, age_at_onset != ".")
mci_baseline
```

```{r}
# count of recruited participant
dim(baseline)[1]
```
```{r}
# count of participant developed MCI
dim(mci_baseline)[1]
```

```{r}
# average age
mean(pull(baseline, current_age))
```
```{r}
# Women APOE4 carriers proportion
dim(filter(baseline, apoe4 == "carrier" & sex == "Female"))[1] / dim(filter(baseline, sex == "Female"))[1]
```


### Biomarker Dataset
```{r}
amyloid <- 
  # load data
  read_csv("./datasets/HW2/data_mci/mci_amyloid.csv", skip = 1) |>
  # clean column name
  janitor::clean_names()  
```
### Data Combination
```{r}
# check whether some participants appear in only the baseline or amyloid datasets
both_id <- intersect(pull(baseline, id), pull(amyloid, study_id))
# baseline
only_baseline <- c()
for(id in pull(baseline, id)){
  if (!id %in% both_id)
    only_baseline <- append(only_baseline, id)
}
# amyloid
only_amyloid <- c()
for(id in pull(amyloid, study_id)){
  if (!id %in% both_id)
    only_amyloid <- append(only_amyloid, id)
}
print("Participant only in baseline:")
only_baseline
print("Participant only in amyloid:")
only_amyloid
```

```{r}
# combine datasets
participants <- inner_join(baseline, amyloid, by = join_by(id == study_id))
participants
```
```{r}
# brief summary of merged dataset
summary(participants)
```

```{r}
# write combined data as csv file
write_csv(participants, "./output/hw2_problem3_output.csv")
```

